---
title: "Replication"
author: "Paul K. Yu"
date: "2024-11-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exploratory data analysis

## Do this once (set eval to TRUE only once)

```{r eda install, eval=FALSE}
# Install required libraries if not already installed
if (!requireNamespace("pheatmap", quietly = TRUE)) install.packages("pheatmap")
if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
```

# Simulate data

```{r}
n <- 100 ## number of subjects
g <- 5000 ## number of gene features
set.seed(123)
epxression <- matrix(rnorm(n*g), nrow=g, ncol=n) ## briefly simulate gene expression data
outcome <- matrix(rnorm(n), nrow=n, ncol=1) ## simulate a continuous outcome variable

# Add generic row and column names
rownames(epxression) <- paste0("Gene", seq_len(g))  # Generic gene names
colnames(epxression) <- paste0("Subject", seq_len(n))  # Generic subject names

rownames(outcome) <- paste0("Subject", seq_len(n))  # Generic subject names
colnames(outcome) <- paste0("Outcome", seq_len(1))  # Generic outcome names

# Verify the structure of the matrix
str(epxression)
str(outcome)

```

# Summary of the dataset

Get summary of gene expression data

```{r}
summary(epxression)
```

Get summary of outcome data

```{r}
summary(outcome)
```

Check for missing values

```{r}
sum(is.na(epxression))
sum(is.na(outcome))
```

# Distribution of the dataset

Get distribution of gene expression data

```{r}
library(reshape2)
library(ggplot2)

df <- melt(epxression)

p <- ggplot(df, aes(x = value)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density of Gene Expression", x = "Expression Level", y = "Density") +
  theme_minimal()

print(p)
```

Get distribution of outcome data

```{r}
df <- as.data.frame(outcome)

p <- ggplot(df, aes(x = outcome)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density of Outcome", x = "Outcome", y = "Density") +
  theme_minimal()

print(p)
```

# Correlation analysis

```{r}
library(pheatmap)

# Compute correlations
cor_matrix <- cor(epxression, use = "pairwise.complete.obs")

# Visualize correlations as a heatmap
pheatmap(cor_matrix, 
         clustering_distance_rows = "correlation", 
         clustering_distance_cols = "correlation",
         main = "Correlation Heatmap of Genes")

```

# Heatmap of Gene Expression

```{r}
# Scale the data (row-wise)
scaled_epxr <- t(scale(t(epxression)))

# Plot the heatmap
pheatmap(scaled_epxr, 
         show_rownames = TRUE, 
         show_colnames = TRUE, 
         cluster_rows = TRUE, 
         cluster_cols = TRUE, 
         main = "Heatmap of Gene Expression")

```

# Replication

## Do this once (set eval to TRUE only once)

```{r install, eval=FALSE}
install.packages("devtools")
library(devtools)

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("qvalue")

install_github("LingsongMeng/GuidedSparseKmeans") 

```

```{r}
library(GuidedSparseKmeans)

# Estimate tuning parameter K and lambda
KLam <- GuidedSparseKmeans.KLam(x=t(epxression), z=outcome, pre.K = NULL, s.one=10, model="linear", nstart = 20, maxiter = 15, silence = F)

# Estimate tuning parameter s
s <- GuidedSparseKmeans.S(x=t(epxression), z=outcome, K=3, s=c(8:12), lam=1, model="linear", nstart = 20, maxiter = 15, nperms = 50, silence = F)

# Obtain gene selection and clustering results
results <- GuidedSparseKmeans(x=t(epxression), outcome, K=3, s=10, lam=1, model="linear", nstart=20, maxiter=15)

## Users can also obtain guidance first, then use the guidance in the next several steps.
# Obtain guidance 
R2 <- getR2(t(epxression), outcome, model="linear")

# Estimate tuning parameter K and lambda
KLam <- GuidedSparseKmeans.KLam.R2out(x=t(epxression), R2.per=R2, pre.K = NULL, s.one=10, nstart = 20, maxiter = 15, silence = F)

# Estimate tuning parameter s
s <- GuidedSparseKmeans.S.R2out(x=t(epxression), R2.per=R2, K=3, s=c(8:12), lam=1, nstart = 20, maxiter = 15, nperms = 50, silence = F)

# Obtain gene selection and clustering results
results <- GuidedSparseKmeans.R2out(x=t(epxression), R2.per=R2, K=3, s=10, lam=1, nstart=20, maxiter=15)

```

# Heatmap of clustered gene expression data

```{r}
# Assign cluster IDs to subjects
subj_clusters <- data.frame(
  Subject = colnames(scaled_epxr), # Subject names
  Cluster = results[[1]]$clusters  # Corresponding cluster IDs
)

# View the mapping
head(subj_clusters)

# Create a column annotation for clusters
annotation_col <- data.frame(
  Cluster = factor(results[[1]]$clusters)  # Convert cluster IDs to factors
)

rownames(annotation_col) <- colnames(scaled_epxr)  # Set rownames to subject names

# Order the matrix by clusters
sorted_indices <- order(annotation_col$Cluster)
epxr_sorted <- scaled_epxr[, sorted_indices]

# Update the annotation to match the sorted order
annotation_col <- annotation_col[sorted_indices, , drop = FALSE]

# Calculate row means (or other statistics like row median)
row_means <- rowMeans(epxr_sorted)

# Sort the row indices based on row means
sorted_indices <- order(row_means, decreasing = TRUE)

# Reorder the expression matrix based on sorted indices
epxr_sorted <- epxr_sorted[sorted_indices, ]

color_palette <- colorRampPalette(c("green", "red"))(100)
breaks <- seq(-4, 4, length.out = 101)

# Generate heatmap with annotation
pheatmap(epxr_sorted, 
         cluster_cols = FALSE,
         cluster_rows = FALSE,
         show_rownames = FALSE,
         show_colnames = FALSE,
         color = color_palette,
         breaks = breaks,
         annotation_col = annotation_col,  # Add column annotations
         main = "Heatmap with Gene Clusters")

```
